Physics of Tidal Dissipation in Early-Type Stars and White Dwarfs: Hydrodynamical Simulations of Internal Gravity Wave Breaking in Stellar Envelopes

### 2020-02-19 writeups/Su_IGW_break.tex
stash WIP Daniel review comments

    \documentclass[
        fleqn,
        usenatbib,
        % referee,
    ]{mnras}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        ae, aecompl,
        graphicx,
        booktabs,
    }
    \usepackage[T1]{fontenc}
    \usepackage[
        labelfont=bf, % caption names are labeled in bold
        font=scriptsize % smaller font for captions
    ]{caption}
    \usepackage[caption=false]{subfig} % subfigures

    \newcommand*{\scinot}[2]{#1\times10^{#2}}
    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\rtd}[2]{\frac{\mathrm{d}^2#1}{\mathrm{d}#2^2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\ptd}[2]{\frac{\partial^2#1}{\partial#2^2}}
    % inline
    \newcommand*{\mdil}[2]{\mathrm{D}#1/\mathrm{D}#2}
    \newcommand*{\pdil}[2]{\partial#1/\partial#2}
    \newcommand*{\rdil}[2]{\mathrm{d}#1/\mathrm{d}#2}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\left\langle#1\right\rangle}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \newcommand*{\bm}[1]{\mathbf{#1}}
    \newcommand*{\uv}[1]{\hat{\mathbf{#1}}}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}
    \DeclareMathOperator*{\med}{med}
    \DeclareMathOperator*{\erf}{erf}

\title[Physics of Tidal Dissipation]{Physics of Tidal Dissipation in Early-Type
Stars and White Dwarfs: Hydrodynamical Simulations of Internal Gravity Wave
Breaking in Stellar Envelopes}
\author[Y. Su et\ al.]{
Yubo Su,$^1$,
Daniel Lecoanet,$^2$
Dong Lai$^{1,3}$
\\
$^1$ Cornell Center for Astrophysics and Planetary Science, Department of
Astronomy, Cornell University, Ithaca, NY 14853, USA
\\
$^2$ Princeton Center for Theoretical Science, Princeton University, Princeton,
NJ 08544, USA
\\
$^3$ Tsung-Dao Lee Institute, Shanghai Jiao Tong University, Shanghai, 200240, China
}

@@

\begin{abstract}
    In binaries composed of early-type stars and white dwarfs, the dominant
    tidal process involves the excitation of internal gravity waves (IGWs) that
    propagate towards the stellar surface and their dissipation via nonlinear
    wave breaking. We perform 2D hydrodynamical simulations of this wave
    breaking process in a stratified isothermal atmosphere. We find that, after
    an initial transient phase, the dissipation of the IGWs naturally generate a
    sharp critical layer, separating the lower stationary region (with no mean
    flow) and the upper ``synchronized'' region (with the mean flow velocity
    equal to the horizontal wave phase speed). We find that the indefinite
    steepening of this critical layer is suppressed by Kelvin-Helmholtz
    instabilities, and the width of the layer is determined by the Richardson
    criterion. We study the absorption and reflection of incident waves off the
    critical layer and provide analytical formulae describing its evolution. The
    result of this study is important for characterizing the evolution of
    tidally heated white dwarfs and other binary stars.
\end{abstract}

@@

The physical processes responsible for tidal evolution in close binaries often
involve the excitation and dissipation of internal waves, going beyond the
``weak friction'' of equilibrium tides \citep[see][for a
review]{ogilvie2014tidal}. In particular, internal gravity waves (IGWs), arising
from buoyancy of stratified stellar fluid, play an important role in several
types of binary systems. In solar-type stars with radiative cores and convective
envelopes, IGWs are excited by tidal forcing at the radiative-convective
boundary and propagate inward; as the wave amplitude grows due to geometric
focusing, nonlinear effects can lead to efficient damping of the wave
\citep{goodman1998dynamical,barker_ogilvie,essick2015orbital}. In early-type
main-sequence stars, with convective cores and radiative envelopes, IGWs are
similarly excited at the convective-radiative interface but travel toward the
stellar surface; nonlinearity develops as the wave amplitude grows, leading to
efficient dissipation \citep{zahn75,zahn77}. As the outgoing wave deposites its
angular momentum to the stellar surface layer, a critical layer may form and the
star is expected to synchronize from outside-in \citep{gn89}.

@@

In this paper, we perform numerical simulations of IGW breaking in a
plane-parallel stratified atmosphere (stellar enevelope). We use the
pseudo-spectral code Dedalus \citep{dedalus,dedalus2} and a 2D Cartesian
geometry, and consider IGWs propagating into an isothermal fluid initially at
rest. We find that, after an initial transient phase, a \emph{critical layer}
naturally develops, separating a lower zone that has no horizontal mean flow
and an upper zone with mean flow at the horizontal phase velocity of the IGW\@.
The major part of our paper is dedicated to characterizing the behavior of the
critical layer when interacting with a continuous train of IGW excited from the
bottom of the atmosphere. IGWs are generally \emph{anti}-diffusive, in that they
steepen shear flows \citep{lindzen_qbo,lecoanet_meanflow} and act to narrow the
critical layer. We find this steepening is suppressed by the Kelvin-Helmholtz
instability and turbulence within the narrow critical layer. By careful
accounting of the momentum flux budget about the critical layer, we are able to
model the reflection and absorption of the incident IGW, and the slow downward
propagation of the critical layer.

While the motivation of our study is to understand tidal dissipation in WD and
early-type stellar binaries, the IGW breaking process studied in this paper is
also quite relevant to the circulation dynamics of planetary atmospheres
\citep[see e.g.][]{lindzen1981turbulence,holton1983influence}.

This paper is organized as follows. In Section~\ref{s:equations} we present the
system of equations used in our simulations. In Section~\ref{s:theory}, we
review the existing understanding of wave breaking and present new analytical
results on IGWs around a critical layer. In Section~\ref{s:numerics} we describe
our numerical setup and in Section~\ref{s:weak_sim} we validate our method in
the weak-forcing limit against linear theory. In Section~\ref{s:sim}, we present
the results of simulations of IGW breaking and our characterization of the
critical layer. We summarize and conclude in Section~\ref{s:discussion}.

@@

We consider a stratified, incompressible, isothermal fluid representing a
stellar envelope or atmosphere. We use barotropic equation of state $P(\rho, T)
= P(\rho)$ as a first approximation. We study dynamics in 2D, so that fluid
variables depend only on the Cartesian coordinates $x$ and $z$. While it is
well-known wave breaking is a 3D process \citep{klostermeyer,winters1994}, the
dynamical effect of the breaking process is likely to be similar in 2D
\citep{barker_ogilvie}. We approximate the gravitational field as uniform,
pointing in the $(-\uv{z})$ direction. The plane-parallel approximation is
justified since wave breaking generally occurs near the stellar surface. The
background density stratification is given by
\begin{equation}
    \overline{\rho} = \overline{\rho}_0 e^{-z/H},
\end{equation}
with $\overline{\rho}_0$ some reference density. We generally denote background
quantities with overbars and perturbation quantities with primes.

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\bm{u}}{t} + \frac{\bm{\nabla}P}{\rho} + g\uv{z} &=
            0\label{eq:nl_mom},
    \end{align}
\end{subequations}
where $\mdil{}{t} = \pdil{}{t}\,+\,\p{\bm{u} \cdot \bm{\nabla}}$ is the
Lagrangian or material derivative, and $\bm{u}, \rho, P$ denote the velocity
field, density and pressure respectively. We denote $-g\uv{z}$ constant
gravitational acceleration. Note that hydrostatic equilibrium ($\pdil{}{t} = 0$)
of the background implies $\bm{\nabla}\overline{P} = -\overline{\rho} g\uv{z}$,
so that $\overline{P} = \overline{\rho} gH$. A vertically-stratified shear flow
$\overline{u}_x(z)\uv{x}$ is permitted in hydrostatic equilibrium, but we will
assume no such background flow, so $\bm{u} = \bm{u}'$. Physically, this
assumption corresponds to a non-rotating star, or going to the corotating frame
of a rigidly rotating star.

For convenience, we introduce the dimensionless density variable $\Upsilon$ and
the reduced pressure $\varpi$ \citep[e.g.][]{lecoanet_anel} via
\begin{align}
    \Upsilon &\equiv \ln \frac{\rho}{\bar{\rho}},\\
    \varpi &\equiv \frac{P}{\rho}.
\end{align}
These variables automatically enforce $\rho > 0$ and eliminate the stiff term
$\vec{\nabla} P / \rho$ in the Euler equation. In terms of $\Upsilon$ and
$\varpi$, the second two equations in~\eqref{se:nl_orig} become
\begin{subequations}\label{se:nl_upsilon}
    \begin{align}
        \md{\Upsilon}{t} + u_z \pd{\ln \overline{\rho}}{z} &= 0
            ,\label{eq:nl_up_density} \\
        \md{\bm{u}}{t} + \bm{\nabla}\varpi + \varpi\bm{\nabla}\Upsilon
            - \frac{\varpi}{H}\uv{z} + g\uv{z} &= 0\label{eq:nl_upsilon_u}.
    \end{align}
\end{subequations}
Hydrostatic equilibrium corresponds to $\Upsilon = 0, \overline{\varpi} = gH$.

@@

In the small perturbation limit, we may linearize Eq.~\eqref{se:nl_upsilon}. The
solution in this linear regime is given by \citep{drazin,sutherland0}
\begin{equation}
    u_z'\p{x, z, t} = Ae^{z/2H}\cos\p{k_{x}x + k_{z}z - \omega t},
        \label{eq:lin_sol}
\end{equation}
where $A$ is a constant amplitude, and the frequency $\omega$ and the wave
number $\p{k_x, k_z}$ satisfy the dispersion relation
\begin{equation}
    \omega^2 = \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}.
        \label{eq:disp_rel}
\end{equation}
Our equations are valid in the limit of large sound speed ($c_s \to \infty$), in
which the \emph{Brunt-V\"ais\"al\"a frequency}, $N$, is given by
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \frac{1}{c_s^2}} = \frac{g}{H},
\end{equation}
is constant. Other dynamical quantities are simply related to $u'_z$.

@@

    \item The averaged horizontal momentum flux $F$ (in the $+\uv{z}$
        direction) carried by the IGW is defined by
        \begin{equation}
            F(z, t) \equiv \ev{\rho u_{x}' u_{z}'}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_{x}'u_{z}'\;\mathrm{d}x.
                    \label{eq:F_def}
        \end{equation}
        The notation $\ev{\dots}_x$ denotes averaging over the $x$ domain. For
        the linear solution (Eq.~\ref{eq:lin_sol}), this evaluates to
        \begin{equation}
            F \approx -\frac{A^2}{2}\overline{\rho}_0\frac{k_{z}}{k_{x}},
                    \label{eq:S_lin}
        \end{equation}
        Thus, indeed $F > 0$ for an upward propagating IGW ($c_{g, z} > 0$).
\end{enumerate}

@@

In the linearized system, the effect of this forcing can be solved exactly
(see Appendix~\ref{s:force_solved}). If we approximate $\abs{k_zH} \gg 1, \sigma
\ll H$, the solution can be approximated as two plane waves propagating away
from the forcing zone
\begin{align}
    u_{z}&(x, z, t) \approx{} \frac{C}{2k_z}\frac{gk_x^2}{\omega^2}
        \exp\p{-\frac{k_z^2\sigma^2}{2}}
        \sqrt{2\pi \sigma^2} \nonumber\\
        &{}\times\begin{cases}
        e^{\frac{z - z_0}{2H}}\sin\p{k_{x}x + k_{z}(z - z_0) - \omega t
            + \frac{k_z\sigma^2}{2H}}
            & \text{for }z > z_0,\\[5pt]
        e^{\frac{z - z_0}{2H}}\sin\p{k_{x}x - k_{z}(z - z_0) - \omega t
            + \frac{k_z\sigma^2}{2H}}
            & \text{for }z < z_0.\\
    \end{cases}\label{eq:uz_lin}
\end{align}
The $z > z_0$ region contains an upward propagating IGW wavetrain. The $x$
component of the velocity can be obtained by the incompressibility constraint
(Eq.~\ref{eq:nl_incomp}).

@@

As the upward propagating IGW grows in amplitude ($\abs{\bm{u}} \propto
e^{z/2H}$), it is expected to break due to nonlinear effects. We can estimate
the height of wave breaking using the condition $\abs{\bm{u}} \sim \omega /
\abs{\bm{k}}$. This can be rewritten using the Lagrangian displacement $\bm{\xi}
= \bm{u} / \p{-i\omega}$:
\begin{equation}
    \abs{\xi_z k_z} \sim 1.\label{eq:nl}
\end{equation}

\citet{drazin, klostermeyer, winters1994} describe the onset of wave breaking in
some detail. At intermediate amplitudes, wave breaking occurs via triadic
resonances, transfering energy from the ``parent’’ IGW to ``daughter’’ waves on
smaller length scales that efficiently damp. The horizontal momentum flux
decreases from $F$ to $0$ over this breaking region. The lost flux is deposited
into a horizontal mean flow
\begin{equation}
    \overline{U}(z, t) \equiv \ev{u_x}_x.\label{eq:mean_flow}
\end{equation}
As the mean flow grows, a \emph{critical layer} may form, as discussed below.

@@

A horizontal shear flow $\overline{U}(z, t)\uv{x}$ enters the fluid equations
via the Lagrangian derivative, which can be decomposed as
\begin{equation}
    \md{}{t} = \pd{}{t} + \overline{U} \pd{}{x} + \p{\bm{u}' \cdot \vec{\nabla}},
\end{equation}
where $\bm{u}'$ is the velocity field \emph{without} the shear flow. Thus,
$\overline{U}$ has the effect of Doppler shifting the time derivative into the
frame comoving with the mean flow. If $\overline{U}$ is roughly constant, then
the behavior of a linear plane-wave perturbation satisfies the modified
dispersion relation
\begin{equation}
    \p{\omega - \overline{U}k_x}^2 =
        \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}.
        \label{eq:disp_rel_U}
\end{equation}
This is just Eq.~\eqref{eq:disp_rel} with $\omega \to \omega - \overline{U}
k_x$. It is apparent that if $\overline{U} = \overline{U}_c$, where
\begin{equation}
    \overline{U}_c \equiv \frac{\omega}{k_x},\label{eq:u_crit}
\end{equation}
then the dispersion relation is singular and the linear solution breaks down.
Physically, this corresponds to the Doppler-shifted frequency of the IGW being
zero. Anywhere $\overline{U} = \overline{U}_c$ is called a \emph{critical
layer}.

@@

We solve Eqs.~\eqref{eq:nl_incomp},~\eqref{eq:nl_upsilon_u},
and~\eqref{eq:vol_drive} in a Cartesian box with size $L_x, L_z$. We choose
periodic boundary conditions in both the $x$ and $z$ direction. We damp
perturbations to zero near the top/bottom of the domain using damping zones
(see Section~\ref{ss:damping}). We expand all variables as Fourier series
with $N_x$ and $N_z$ modes, and use the $3/2$ dealiasing rule to avoid aliasing
errors in the nonlinear terms \citep{boyd}.

@@

    \item $C$: In our weak forcing simulations, we first choose the forcing
        strength $C$ (see Eq.~\eqref{eq:vol_drive}) such that $\abs{\xi_z k_z}
        \ll 1$ is satisfied everywhere in the simulation domain. This constrains
        $C$ by Eq.~\eqref{eq:uz_lin}.

    \item $\nu$: Nonlinear effects transfer wave energy from the injection
        wavenumber $\bm{k}$ to larger wavenumbers. Our spectral method does not
        have any numerical viscosity, so diffusivity must be introduced into the
        equations to regularize the systems at large wavenumbers. We add
        viscosity and diffusivity to the system in a way that conserves
        horizontal momentum (see Appendix~\ref{se:strat_impl} for details). To
        quantify $\nu$, we define the dimensionless Reynolds number
        \begin{equation}
            \mathrm{Re} \equiv \frac{\omega}{\nu k_{z}^2}
                = \frac{\omega}{\nu}\p{\frac{H}{2\pi}}^2.\label{eq:re_def}
        \end{equation}
        We use $\mathrm{Re} \gg 1$ in our simulations\footnote{This condition
        is always satisfied in stars. For example, in WDs, the dominant linear
        dissipation mechanism of g-modes is radiative damping, with damping rate
        ranging from $10^{-11}$--$10^-4$ of the mode frequency \citep{fullerI}.
        This corresponds to a small effective viscosity or $\mathrm{Re} \gg
        1$.}. Note that $\nu = 0$ changes the order of the PDE system, as the
        highest-order derivatives appear only in the dissipitive terms, so only
        nonzero $\nu$ can be used even in weakly forced simulations.
\end{enumerate}

>>>>

% TODO
% - Figure 1: Could you include a tick mark at t=0? Does the curve start at t=0?
% Section 6.2 needs more narrative discussion of the results. We should add another paragraph saying that the typical Richardson number of the critical layer is between 0.25 and 0.5, characteristic of Kelvin-Helmholtz instabilities. This suggestions the critical layer width is self-consistently determined by the ``turbulence’’ in our simulations. The layer width does not change with resolution. We should also say that here we calculate Ri using the background buoyancy frequency, because the fluctuations do not change N^2 significantly; thus, Ri~0.5 can be used to solve for the critical layer width in a star, as we know N^2 and U_c (corresponding to the tidal frequency Omega).
% rewrite eq_fadef in explicit integrals
% Investigate a smarter time averaging for the "average over everything
% predictor"?
    \documentclass[
        fleqn,
        usenatbib,
        % referee,
    ]{mnras}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        ae, aecompl,
        graphicx,
        booktabs,
    }
    \usepackage[T1]{fontenc}
    \usepackage[
        labelfont=bf, % caption names are labeled in bold
        font=scriptsize % smaller font for captions
    ]{caption}
    \usepackage[caption=false]{subfig} % subfigures

    \newcommand*{\scinot}[2]{#1\times10^{#2}}
    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\rtd}[2]{\frac{\mathrm{d}^2#1}{\mathrm{d}#2^2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\ptd}[2]{\frac{\partial^2#1}{\partial#2^2}}
    % inline
    \newcommand*{\mdil}[2]{\mathrm{D}#1/\mathrm{D}#2}
    \newcommand*{\pdil}[2]{\partial#1/\partial#2}
    \newcommand*{\rdil}[2]{\mathrm{d}#1/\mathrm{d}#2}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\left\langle#1\right\rangle}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \newcommand*{\bm}[1]{\mathbf{#1}}
    \newcommand*{\uv}[1]{\hat{\mathbf{#1}}}
    \DeclareMathOperator*{\med}{med}
    \DeclareMathOperator*{\erf}{erf}

\title[Physics of Tidal Dissipation]{Physics of Tidal Dissipation in Early-Type
Stars and White Dwarfs: Hydrodynamical Simulations of Internal Gravity Wave
Breaking in Stellar Envelopes}
\author[Y. Su et\ al.]{
Yubo Su,$^1$,
Daniel Lecoanet,$^2$
Dong Lai$^{1,3}$
\\
$^1$ Cornell Center for Astrophysics and Planetary Science, Department of
Astronomy, Cornell University, Ithaca, NY 14853, USA
\\
$^2$ Princeton Center for Theoretical Science, Princeton University, Princeton,
NJ 08544, USA
\\
$^3$ Department of Astrophysical Sciences, Princeton University, Princeton NJ
08544, USA
\\
$^4$ Tsung-Dao Lee Institute, Shanghai Jiao Tong University, Shanghai, 200240, China
}

@@

\begin{abstract}
    In binaries composed of early-type stars and white dwarfs, the dominant
    tidal process involves the excitation of internal gravity waves (IGWs),
    which propagate towards the stellar surface, and their dissipation via
    nonlinear wave breaking. We perform 2D hydrodynamical simulations of this
    wave breaking process in a stratified isothermal atmosphere. We find that,
    after an initial transient phase, the dissipation of the IGWs naturally
    generate a sharp critical layer, separating the lower stationary region
    (with no mean flow) and the upper ``synchronized'' region (with the mean
    flow velocity equal to the horizontal wave phase speed). While the critical
    layer is steepened by these waves, it is simultaneously broadened by
    Kelvin-Helmholtz instabilities such that, in steady state, the critical
    layer width is determined by the Richardson criterion. We study the
    absorption and reflection of incident waves off the critical layer and
    provide analytical formulae describing its evolution. The result of this
    study is important for characterizing the evolution of tidally heated white
    dwarfs and other binary stars.
\end{abstract}

@@

The physical processes responsible for tidal evolution in close binaries often
involve the excitation and dissipation of internal waves, going beyond the
``weak friction'' of equilibrium tides \citep[see][for a
review]{ogilvie2014tidal}. In particular, internal gravity waves (IGWs), arising
from buoyancy of stratified stellar fluid, play an important role in several
types of binary systems. In solar-type stars with radiative cores and convective
envelopes, IGWs are excited by tidal forcing at the radiative-convective
boundary and propagate inward; as the wave amplitude grows due to geometric
focusing, nonlinear effects can lead to efficient damping of the wave
\citep{goodman1998dynamical,barker_ogilvie,essick2015orbital}. In early-type
main-sequence stars, with convective cores and radiative envelopes, IGWs are
similarly excited at the convective-radiative interface but travel toward the
stellar surface; nonlinearity develops as the wave amplitude grows, leading to
efficient dissipation \citep{zahn75,zahn77}. As the outgoing wave deposits its
angular momentum to the stellar surface layer, a critical layer may form and the
star is expected to synchronize from outside-in \citep{gn89}.

@@

In this paper, we perform numerical simulations of IGW breaking in a
plane-parallel stratified atmosphere (a simple model for a stellar envelope). We
use the pseudo-spectral code Dedalus \citep{dedalus,dedalus2} and a 2D Cartesian
geometry, and consider IGWs propagating into an isothermal fluid initially at
rest. We find that, after an initial transient phase, a \emph{critical layer}
naturally develops, separating a lower zone that has no horizontal mean flow and
an upper zone with mean flow at the horizontal phase velocity of the IGW\@. The
major part of our paper is dedicated to characterizing the behavior of the
critical layer when interacting with a continuous train of IGW excited from the
bottom of the atmosphere. IGWs are generally \emph{anti}-diffusive, in that they
steepen shear flows \citep{lindzen_qbo,lecoanet_meanflow} and act to narrow the
critical layer. We find this steepening is counter-balanced by the
Kelvin-Helmholtz instability and turbulence within the narrow critical layer. By
careful accounting of the momentum flux budget about the critical layer, we are
able to model the reflection and absorption of the incident IGW, and the slow
downward propagation of the critical layer.

While the motivation of our study is to understand tidal dissipation in WD and
early-type stellar binaries, the IGW breaking process studied in this paper is
also quite relevant to the circulation dynamics of planetary atmospheres
\citep[see e.g.][]{lindzen1981turbulence,holton1983influence,baldwin2001quasi}.

This paper is organized as follows. In Section~\ref{s:equations} we present the
system of equations used in our simulations. In Section~\ref{s:theory}, we
review the existing understanding of wave breaking and present analytical
results characterizing IGW behavior near a critical layer. In
Section~\ref{s:numerics} we describe our numerical setup and in
Section~\ref{s:weak_sim} we validate our method in the weak-forcing limit
against linear theory. In Section~\ref{s:sim}, we present the results of
simulations of IGW breaking and our characterization of the critical layer. We
summarize and conclude in Section~\ref{s:discussion}.

@@

We consider a stratified, incompressible, isothermal fluid representing a
stellar envelope or atmosphere. We approximate the equation of state as
baratropic, i.e.\ $P(\rho, T) = P(\rho)$. We study dynamics in 2D, so that fluid
variables depend only on the Cartesian coordinates $x$ and $z$. While it is
well-known wave breaking is a 3D process \citep{klostermeyer,winters1994}, the
dynamical effect of the breaking process is likely to be similar in 2D
\citep{barker_ogilvie}. We approximate the gravitational field as uniform,
pointing in the $(-\uv{z})$ direction. The plane-parallel approximation is
justified since wave breaking generally occurs near the stellar surface. The
background density stratification is given by
\begin{equation}
    \overline{\rho} = \overline{\rho}_0 e^{-z/H},
\end{equation}
with $\overline{\rho}_0$ some reference density. We denote background quantities
with overbars and perturbation quantities with primes.

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\bm{u}}{t} + \frac{\bm{\nabla}P}{\rho} + g\uv{z} &=
            0\label{eq:nl_mom},
    \end{align}
\end{subequations}
where $\mdil{}{t} = \pdil{}{t}\,+\,\p{\bm{u} \cdot \bm{\nabla}}$ is the
Lagrangian or material derivative, and $\bm{u}, \rho, P$ denote the velocity
field, density and pressure respectively. The constant gravitational
acceleration is $-g\uv{z}$. For this isothermal background, hydrostatic
equilibrium implies $P = \overline{\rho} g H$. We assume there is initially no
background flow, so $\bm{u} = \bm{u}'$. Physically, this assumption corresponds
to a non-rotating star.

For convenience, we introduce the dimensionless density variable $\Upsilon$ and
the reduced pressure $\varpi$ \citep[e.g.][]{lecoanet_anel} via
\begin{align}
    \Upsilon &\equiv \ln \frac{\rho}{\bar{\rho}},\\
    \varpi &\equiv \frac{P}{\rho}.
\end{align}
These variables automatically enforce $\rho > 0$ and eliminate the stiff term
$\bm{\nabla} P / \rho$ in the Euler equation. In terms of $\Upsilon$ and
$\varpi$, the second two equations in~\eqref{se:nl_orig} become
\begin{subequations}\label{se:nl_upsilon}
    \begin{align}
        \md{\Upsilon}{t} + u_z \pd{\ln \overline{\rho}}{z} &= 0
            ,\label{eq:nl_up_density} \\
        \md{\bm{u}}{t} + \bm{\nabla}\varpi + \varpi\bm{\nabla}\Upsilon
            - \frac{\varpi}{H}\uv{z} + g\uv{z} &= 0\label{eq:nl_upsilon_u}.
    \end{align}
\end{subequations}
Hydrostatic equilibrium corresponds to $\Upsilon = 0, \overline{\varpi} = gH$.

@@

In the small perturbation limit, we may linearize Eq.~\eqref{se:nl_upsilon}. The
solution in this linear regime is given by \citep{drazin,sutherland0}
\begin{equation}
    u_z'\p{x, z, t} = Ae^{z/2H}\cos\p{k_{x}x + k_{z}z - \omega t},
        \label{eq:lin_sol}
\end{equation}
where $A$ is a constant amplitude, and the frequency $\omega$ and the wave
number $\p{k_x, k_z}$ satisfy the dispersion relation
\begin{equation}
    \omega^2 = \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}.
        \label{eq:disp_rel}
\end{equation}
Our equations are valid in the limit of large sound speed ($c_s \to \infty$), in
which the \emph{Brunt-V\"ais\"al\"a frequency}, $N$, is given by
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \frac{1}{c_s^2}} = \frac{g}{H},
\end{equation}
and is constant. Other dynamical quantities are simply related to $u'_z$.

@@

    \item The averaged horizontal momentum flux $F$ (in the $+\uv{z}$
        direction) carried by the IGW is defined by
        \begin{equation}
            F(z, t) \equiv \ev{\rho u_{x}' u_{z}'}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_{x}'u_{z}'\;\mathrm{d}x.
                    \label{eq:F_def}
        \end{equation}
        The notation $\ev{\dots}_x$ denotes averaging over the $x$ direction.
        For the linear solution (Eq.~\ref{eq:lin_sol}), this evaluates to
        \begin{equation}
            F \approx -\frac{A^2}{2}\overline{\rho}_0\frac{k_{z}}{k_{x}},
                    \label{eq:S_lin}
        \end{equation}
        Thus, indeed $F > 0$ for an upward propagating IGW ($c_{g, z} > 0$).
\end{enumerate}

@@

In the linearized system, the effect of this forcing can be solved exactly
(see Appendix~\ref{s:force_solved}). In the limit $\abs{k_zH} \gg 1, \sigma
\ll H$, the solution can be approximated as two plane waves propagating away
from the forcing zone
\begin{align}
    u_{z}&(x, z, t) \approx{} \frac{C}{2k_z}\frac{gk_x^2}{\omega^2}
        \exp\p{-\frac{k_z^2\sigma^2}{2}}
        \sqrt{2\pi \sigma^2} \nonumber\\
        &{}\times\begin{cases}
        e^{\frac{z - z_0}{2H}}\sin\p{k_{x}x + k_{z}(z - z_0) - \omega t
            + \frac{k_z\sigma^2}{2H}}
            & \text{for }z > z_0,\\[5pt]
        e^{\frac{z - z_0}{2H}}\sin\p{k_{x}x - k_{z}(z - z_0) - \omega t
            + \frac{k_z\sigma^2}{2H}}
            & \text{for }z < z_0.\\
    \end{cases}\label{eq:uz_lin}
\end{align}
The $z > z_0$ region contains an upward propagating IGW wavetrain. The $x$
component of the velocity can be obtained by the incompressibility constraint
(Eq.~\ref{eq:nl_incomp}).

@@

As the upward propagating IGW grows in amplitude ($\abs{\bm{u}} \propto
e^{z/2H}$), it is expected to break due to nonlinear effects. We can estimate
the height of wave breaking using the condition $\abs{\bm{u}} \sim \omega /
\abs{\bm{k}}$. This can be rewritten using the Lagrangian displacement
$\boldsymbol{\xi} = \bm{u} / \p{-i\omega}$:
\begin{equation}
    \abs{\xi_z k_z} \sim 1.\label{eq:nl}
\end{equation}

\citet{drazin, klostermeyer, winters1994} describe the onset of wave breaking in
some detail. At intermediate amplitudes, wave breaking occurs via triadic
resonances, transfering energy from the ``parent'' IGW to ``daughter'' waves on
smaller length scales that efficiently damp. The horizontal momentum flux
decreases from $F$ to $0$ over this breaking region. The lost flux is deposited
into a horizontal mean flow
\begin{equation}
    \overline{U}(z, t) \equiv \ev{u_x}_x.\label{eq:mean_flow}
\end{equation}
As the mean flow grows, a \emph{critical layer} may form, as discussed below.

@@

A horizontal shear flow $\overline{U}(z, t)\uv{x}$ enters the fluid equations
via the Lagrangian derivative, which can be decomposed as
\begin{equation}
    \md{}{t} = \pd{}{t} + \overline{U} \pd{}{x} + \p{\bm{u}' \cdot \bm{\nabla}},
\end{equation}
where $\bm{u}'$ is the velocity field \emph{without} the shear flow. Thus,
$\overline{U}$ has the effect of Doppler shifting the time derivative into the
frame comoving with the mean flow. If $\overline{U}$ is roughly constant, then
the behavior of a linear plane-wave perturbation satisfies the modified
dispersion relation
\begin{equation}
    \p{\omega - \overline{U}k_x}^2 =
        \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}.
        \label{eq:disp_rel_U}
\end{equation}
This is just Eq.~\eqref{eq:disp_rel} with $\omega \to \omega - \overline{U}
k_x$. It is apparent that if $\overline{U} = \overline{U}_c$, where
\begin{equation}
    \overline{U}_c \equiv \frac{\omega}{k_x},\label{eq:u_crit}
\end{equation}
then the dispersion relation is singular and the linear solution breaks down.
Physically, this corresponds to the Doppler-shifted frequency of the IGW being
zero. Anywhere $\overline{U} = \overline{U}_c$ is called a \emph{critical
layer}.

@@

We solve Eqs.~\eqref{eq:nl_incomp},~\eqref{eq:nl_upsilon_u},
and~\eqref{eq:vol_drive} in a Cartesian box with size $L_x, L_z$. We choose
periodic boundary conditions in both the $x$ and $z$ direction. To mimic the
absence of physical boundaries at the top/bottom of the simulation domain, we damp
perturbations to zero near the top/bottom using damping zones
(see Section~\ref{ss:damping}). We expand all variables as Fourier series
with $N_x$ and $N_z$ modes, and use the $3/2$ dealiasing rule to avoid aliasing
errors in the nonlinear terms \citep{boyd}.

@@

    \item $C$: In our weak forcing simulations (Section~\ref{s:weak_sim}),
        we first choose the forcing strength $C$ (see Eq.~\eqref{eq:vol_drive})
        to be sufficiently weak such that $\abs{\xi_z k_z} \ll 1$ is satisfied
        everywhere in the simulation domain. This constrains $C$ by
        Eq.~\eqref{eq:uz_lin}. In our wave breaking simulations
        (Section~\ref{s:sim}), we choose larger $C$.

    \item $\nu$: Nonlinear effects transfer wave energy from the injection
        wavenumber $\bm{k}$ to larger wavenumbers. Our spectral method does not
        have any numerical viscosity, so diffusivity must be introduced into the
        equations to regularize the systems at large wavenumbers. We add
        viscosity and diffusivity to the system in a way that conserves
        horizontal momentum (see Appendix~\ref{se:strat_impl} for details). We
        define the dimensionless Reynolds number
        \begin{equation}
            \mathrm{Re} \equiv \frac{\omega}{\nu k_{z}^2}
                = \frac{\omega}{\nu}\p{\frac{H}{2\pi}}^2.\label{eq:re_def}
        \end{equation}
        We use $\mathrm{Re} \gg 1$ in our simulations\footnote{This condition
        is always satisfied in stars. For example, in WDs, the dominant linear
        dissipation mechanism of g-modes is radiative damping, with damping rate
        ranging from $10^{-11}$--$10^{-4}$ of the mode frequency
        \citep{fullerI}. This corresponds to a small effective viscosity or
        $\mathrm{Re} \gg 1$.}.
\end{enumerate}
